---
title: "R Notebook for Intro to R Presentation"
subtitle: "By Cooper and Cecilia on October 14th, 2025" 
output: html_notebook
---

## Silde 3: Motivation & Context
## Slide 4: Reproducibility & Integration

Show how basic R and R Markdown work by exploring the preloaded dataset mtcars

### Steps:

### 1. Open R Markdown

### 2. View the dataset

```{r}
mtcars
```

### 3. Get a summary of the dataset

```{r}
summary(mtcars)
```

### 4. Show a simple plot

```{r}
plot(mtcars$mpg, mtcars$hp)
```

## Silde 5-6: Reproducibility & Integration continued
## Silde 7: Who has used R before?
## Silde 8-10: R Basics
## Silde 11-13: Data Visualization

Show how basic graphs can be made using R

### Steps:

### 1. Create a Scatter Plot

```{r}
plot(mtcars$mpg, mtcars$hp,
    main = "MPG vs Horsepower",
    xlab = "Miles per Gallon",
    ylab = "Horsepower",
    col = "blue",
    pch = 19)
```

### 2. Create a Histogram

```{r}
hist(mtcars$mpg,
    main = "Distribution of MPG",
    xlab = "Miles per Gallon",
    col = "lightgreen",
    border = "black")
```

### 3. Use ggplot2

```{r}
library(ggplot2)
ggplot(mtcars, aes(x = mpg, y = hp)) + geom_point()
```

## Tidyverse Example Workflow

### Steps:

#### 1. Load necessary packages 

```{r}
library(tidyverse)
library(palmerpenguins) #example dataset built in r
```

#### 2. Inspect data

```{r}
str(penguins) 
```

#### 3. Tidy data using pipe operator

Using pipe operator to remove missing values

```{r}
penguins_long <- penguins %>%
  drop_na() %>%   # Remove rows with missing values
  pivot_longer(
    cols = c(bill_length_mm, bill_depth_mm),
    names_to = "measurement",
    values_to = "value"
  )
penguins_long
```

#### 4. Filter data using dplyr

```{r}
penguins_filtered <- penguins_long %>%
  filter(value > 40)
```

#### 5. Plot using ggplot2

Plot bill length by penguin species

```{r}
ggplot(penguins_clean, aes(x = species, y = bill_length_mm, fill = species)) +
  geom_boxplot() +
  labs(
    title = "Bill Length by Penguin Species",
    x = "Species",
    y = "Bill Length (mm)"
  )
```
