---
title: "DSE 511: R Notebook for Intro to R Presentation"
subtitle: "By Cooper and Cecilia on October 14th, 2025" 
output: html_notebook
---

## Silde 3: Motivation & Context

## Slide 4: Reproducibility & Integration

Show how basic R and R Markdown work by exploring the pre-loaded dataset mtcars

### Steps:

### 1. Open R Markdown

### 2. View the dataset

```{r}
mtcars
```

### 3. Using the helper function

```{r}
?summary
```

### 4. Get a summary of the dataset

```{r}
summary(mtcars)
```

## Slide 11-13: Data Visualization

Show how basic graphs can be made using R

### Steps:

### 1. Create a Scatter Plot

```{r}
plot(mtcars$mpg, mtcars$hp,
    main = "MPG vs Horsepower",
    xlab = "Miles per Gallon",
    ylab = "Horsepower",
    col = "blue",
    pch = 19)
```

### 2. Create a Histogram

```{r}
hist(mtcars$mpg,
    main = "Distribution of MPG",
    xlab = "Miles per Gallon",
    col = "lightgreen",
    border = "black")
```

### 3. Use ggplot2

```{r}
library(ggplot2)

ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point(
    aes(color = factor(cyl), size = wt),
    alpha = 0.8
  ) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  scale_color_brewer(palette = "Set2", name = "Cylinders") +
  scale_size_continuous(name = "Weight") +
  labs(
    title = "Relationship Between Miles per Gallon and Horsepower",
    subtitle = "Colored by cylinder count and sized by car weight",
    x = "Miles per Gallon (mpg)",
    y = "Horsepower (hp)",
    caption = "Data: mtcars"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 13, color = "gray30"),
    plot.caption = element_text(size = 10, color = "gray40"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray90"),
    legend.position = "right"
  )
```

## Slides 16-18: Tidyverse Example Workflow

### Steps:

#### 1. Load necessary packages

```{r}
library(tidyverse)
#palmerpenguins - example dataset built in r
```

#### 2. Inspect data

```{r}
str(penguins) 
head(penguins)
```

#### 3. Tidy data using pipe operator

Using pipe operator to remove missing values

```{r}

penguins_clean <- penguins %>%
  mutate(
    bill_ratio = bill_len / bill_dep  # create a new variable
  ) %>%
  drop_na(bill_len, bill_dep, bill_ratio)  # removes rows with NAs

penguins_clean
```

#### 4. Filter data using dplyr

```{r}
penguins_filtered <- penguins_clean %>%
  filter(bill_length_mm > 30)
```

#### 5. Plot using ggplot2

Plot bill length by penguin species

```{r}
ggplot(penguins_filtered, aes(x = species, y = bill_length_mm, fill = species)) +
  geom_boxplot() +
  labs(
    title = "Bill Length by Penguin Species",
    x = "Species",
    y = "Bill Length (mm)"
  )
```

## Slides 19-21: Statistics

In the following code will demonstrate some statistical and modeling capabilities in R

### Generating Data of a Given Distribution

1.  Generate random (r-) Normally-distributed data (-norm)

```{r}
set.seed(123)  # for reproducibility

x <- rnorm(1000, mean = 0, sd = 1)
```

2.  Overlay the theoretical density (d-) curve to the Normal distribution (-norm)

```{r}
# Histogram of data
hist(x, breaks = 30, probability = TRUE,
     main = "Normal Distribution", col = "lightblue")

# Density function overlay
curve(dnorm(x, mean = 0, sd = 1), add = TRUE, col = "red", lwd = 2)
```

3.  Determine the probability that X ≤ 1.96 (pnorm = inverse of qnorm)

```{r}
pnorm(1.96, mean = 0, sd = 1)
```

4.  Find the 97.5% quantile (qnorm = inverse of pnorm)

```{r}
qnorm(0.975, mean = 0, sd = 1)
```
